Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2011-06-02 talk/icooolps2009/talk/talk.tex
Initial draft of a blog post on the survey results.

\date{ICOOOLPS 2009 XXX}

@@

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item idea from Dynamo project: dynamic rewriting of machine code
    \item later used for a lightweight Java JIT
    \item seems to also work for dynamic languages
    \item incorporated into Mozilla's JavaScript VM ("TraceMonkey")
    \end{itemize}
    \pause
    \begin{block}{Basic Assumption of a Tracing JIT}
        \begin{itemize}
        \item programs spend most of their time executing loops
        \item several iterations of a loop are likely to take similar code paths
        \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{(Dis-)Advantages of Tracing JITs}
    \begin{block}{Good Points of the Approach}
        \begin{itemize}
        \item easy and fast machine code generation: needs so support only one path
        \item (things are more complex, but let's ignore that for now)
        \item interpreter does a lot of the work
        \item can be added to an existing interpreter unobtrusively
        \item automatic inlining
        \item produces comparatively little code
        \end{itemize}
    \end{block}
    \pause
    \begin{block}{Bad Points of the Approach}
        \begin{itemize}
        \item unclear whether assumptions are true often enough
        \item switching between interpretation and machine code execution takes time
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: consecutive iterations of the dispatch loop will usually take very different code paths
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter run on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
  \begin{verbatim}
def interpret(bytecode, a):
    regs = [0] * 256
    pc = 0
    while True:
        opcode = ord(bytecode[pc])
        pc += 1
        if opcode == JUMP_IF_A:
            target = ord(bytecode[pc])
            pc += 1
            if a:
                pc = target
        elif opcode == MOV_A_R:
            n = ord(bytecode[pc])
            pc += 1
            regs[n] = a
        elif opcode == MOV_R_A:
            n = ord(bytecode[pc])
            pc += 1
            a = regs[n]
        elif opcode == ADD_R_TO_A:
            n = ord(bytecode[pc])
            pc += 1
            a += regs[n]
        elif opcode == DECR_A:
            a -= 1
        elif opcode == RETURN_A:
            return a
  \end{verbatim}
}

\frame[containsverbatim, plain, shrink=10]{
  \begin{verbatim}
def interpret(bytecode, a):             |
    regs = [0] * 256                    |  # Example bytecode
    pc = 0                              |  # Square the accumulator:
    while True:                         |
        opcode = ord(bytecode[pc])      |  MOV_A_R     0   # i = a
        pc += 1                         |  MOV_A_R     1   # copy of 'a'
        if opcode == JUMP_IF_A:         |
            target = ord(bytecode[pc])  |  # 4:
            pc += 1                     |  MOV_R_A     0   # i--
            if a:                       |  DECR_A
                pc = target             |  MOV_A_R     0
        elif opcode == MOV_A_R:         |
            n = ord(bytecode[pc])       |  MOV_R_A     2   # res += a
            pc += 1                     |  ADD_R_TO_A  1
            regs[n] = a                 |  MOV_A_R     2
        elif opcode == MOV_R_A:         |
            n = ord(bytecode[pc])       |  MOV_R_A     0   # if i!=0:
            pc += 1                     |  JUMP_IF_A   4   #    goto 4
            a = regs[n]                 |
        elif opcode == ADD_R_TO_A:      |  MOV_R_A     2   # return res
            n = ord(bytecode[pc])       |  RETURN_A
            pc += 1                     |
            a += regs[n]                |
        elif opcode == DECR_A:          |
            a -= 1                      |
        elif opcode == RETURN_A:        |
            return a                    |
  \end{verbatim}
}

@@

\begin{frame}
    \frametitle{Idea for a Solution}
    \begin{itemize}
    \item goal: try to trace the loops \emph{in the user program,}
          and not just one iteration of the bytecode dispatch loop
    \item tracing interpreter needs information about the language interpreter
    \item provided by adding \emph{three hints} to the language interpreter
    \end{itemize}
    \pause
    \begin{block}{Hints Give Information About:}
    \begin{itemize}
    \item which variables make up the program counter of the language interpreter
    \item where the bytecode dispatch loop is
    \item which bytecodes can correspond to backward jumps
    \end{itemize}
    \end{block}
\end{frame}

@@

\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE}}
\begin{verbatim}
loop_start(a0, regs0, bytecode0, pc0)
# MOV_R_A 0
opcode0 = strgetitem(bytecode0, pc0)
pc1 = int_add(pc0, Const(1))
guard_value(opcode0, Const(2))
n1 = strgetitem(bytecode0, pc1)
pc2 = int_add(pc1, Const(1))
a1 = list_getitem(regs0, n1)
# DECR_A
...
# MOV_A_R 0
...
# MOV_R_A 2
...
# ADD_R_TO_A 1
...
# MOV_A_R 2
...
# MOV_R_A 0
...
# JUMP_IF_A 4
opcode6 = strgetitem(bytecode0, pc13)
pc14 = int_add(pc13, Const(1))
guard_value(opcode6, Const(3))
target0 = strgetitem(bytecode0, pc14)
pc15 = int_add(pc14, Const(1))
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0, bytecode0, target0)
\end{verbatim}
}

@@

\begin{frame}
    \frametitle{Scaling to Large Interpreters?}
    \begin{itemize}
    \item we can apply this approach to PyPy's Python interpreter (70 KLOC)
    \item speed-ups promising: factor of 6 faster for simple loops with arithmetic
    \item no Python-specific bugs!
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Backup Slides}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
    \frametitle{Timings for Python Interpreter}
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
\begin{block}{Timings}
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}
\end{block}
}

>>>>

\date{6th of July 2009, ICOOOLPS '09, Genova}

@@

\begin{frame}
    \frametitle{Tracing JIT Compilers}
    \begin{itemize}
    \item idea from Dynamo project: dynamic rewriting of machine code
    \item later used for a lightweight Java JIT
    \item seems to also work for dynamic languages (see TraceMonkey)
    \end{itemize}
    \pause
    \begin{block}{Basic Assumption of a Tracing JIT}
        \begin{itemize}
        \item programs spend most of their time executing loops
        \item several iterations of a loop are likely to take similar code paths
        \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{(Dis-)Advantages of Tracing JITs}
    \begin{block}{Good Points of the Approach}
        \begin{itemize}
        \item easy and fast machine code generation: needs so support only one path
        \item (things are more complex, but let's ignore that for now)
        \item interpreter does a lot of the work
        \item can be added to an existing interpreter unobtrusively
        \item automatic inlining
        \item produces comparatively little machine code
        \end{itemize}
    \end{block}
    \pause
    \begin{block}{Bad Points of the Approach}
        \begin{itemize}
        \item unclear whether assumptions are true often enough
        \item switching between interpretation and machine code execution takes time
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Applying a Tracing JIT to an Interpreter}
    \begin{itemize}
    \item Question: What happens if the program is itself a bytecode interpreter?
    \item the (most important) hot loop of a bytecode interpreter is the bytecode dispatch loop
    \item Assumption violated: consecutive iterations of the dispatch loop will usually take very different code paths
    \item what can we do?
    \end{itemize}
    \pause
    \begin{block}{Terminology}
        \begin{itemize}
        \item \emph{tracing interpreter:} the interpreter that originally runs the program and produces traces
        \item \emph{language interpreter:} the bytecode interpreter run on top
        \item \emph{user program:} the program run by the language interpreter
        \end{itemize}
    \end{block}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
  \begin{verbatim}
def interpret(bytecode, a):
    regs = [0] * 256
    pc = 0
    while True:
        opcode = ord(bytecode[pc])
        pc += 1
        if opcode == JUMP_IF_A:
            target = ord(bytecode[pc])
            pc += 1
            if a:
                pc = target
        elif opcode == MOV_R_A:
            n = ord(bytecode[pc])
            pc += 1
            a = regs[n]
        elif opcode == MOV_A_R:
            ...
        elif opcode == ADD_R_TO_A:
            ...
        elif opcode == DECR_A:
            a -= 1
        elif opcode == RETURN_A:
            return a

\frame[containsverbatim, plain, shrink=10]{
  \begin{verbatim}
def interpret(bytecode, a):
    regs = [0] * 256                    |
    pc = 0                              |  # Example bytecode
    while True:                         |  # Square the accumulator:
        opcode = ord(bytecode[pc])      |
        pc += 1                         |  MOV_A_R     0  # i = a
        if opcode == JUMP_IF_A:         |  MOV_A_R     1  # copy of 'a'
            target = ord(bytecode[pc])  |
            pc += 1                     |  # 4:
            if a:                       |  MOV_R_A     0  # i--
                pc = target             |  DECR_A
        elif opcode == MOV_R_A:         |  MOV_A_R     0
            n = ord(bytecode[pc])       |
            pc += 1                     |  MOV_R_A     2  # res += a
            a = regs[n]                 |  ADD_R_TO_A  1
        elif opcode == MOV_A_R:         |  MOV_A_R     2
            ...                         |
        elif opcode == ADD_R_TO_A:      |  MOV_R_A     0  # if i!=0:
            ...                         |  JUMP_IF_A   4  #    goto 4
        elif opcode == DECR_A:          |
            a -= 1                      |  MOV_R_A     2  # return res
        elif opcode == RETURN_A:        |  RETURN_A                     
            return a                    |

@@

\begin{frame}
    \frametitle{Idea for a Solution}
    \begin{itemize}
    \item goal: try to trace the loops \emph{in the user program,}
          and not just one iteration of the bytecode dispatch loop
    \item tracing interpreter needs information about the language interpreter
    \item provided by adding \emph{three hints} to the language interpreter
    \end{itemize}
    \pause
    \begin{block}{Hints Give Information About:}
    \begin{itemize}
    \item which variables make up the program counter of the language interpreter (together those are called \emph{position key})
    \item where the bytecode dispatch loop is
    \item which bytecodes can correspond to backward jumps
    \end{itemize}
    \end{block}
\end{frame}

@@

\begin{frame}
    \frametitle{Modifying Tracing}
    \begin{itemize}
    \item goal: try to trace the loops \emph{in the user program,}
          and not just one iteration of the bytecode dispatch loop
    \item tracing interpreter stops tracing only when:
        \begin{itemize}
        \item it sees a backward jump in the language interpreter
        \item the position key of the language interpreter matches an ealier value
        \end{itemize}
    \item in this way, full user loops are traced
    \end{itemize}
\end{frame}
\frame[containsverbatim, plain, shrink=20]{
    \frametitle{Result When Tracing \texttt{SQUARE}}
\begin{verbatim}
loop_start(a0, regs0, bytecode0, pc0)
# MOV_R_A 0
opcode0 = strgetitem(bytecode0, pc0)
pc1 = int_add(pc0, Const(1))
guard_value(opcode0, Const(2))
n1 = strgetitem(bytecode0, pc1)
pc2 = int_add(pc1, Const(1))
a1 = list_getitem(regs0, n1)
# DECR_A
# MOV_A_R 0
# MOV_R_A 2
# ADD_R_TO_A 1
# MOV_A_R 2
# MOV_R_A 0
...
# JUMP_IF_A 4
opcode6 = strgetitem(bytecode0, pc13)
pc14 = int_add(pc13, Const(1))
guard_value(opcode6, Const(3))
target0 = strgetitem(bytecode0, pc14)
pc15 = int_add(pc14, Const(1))
i1 = int_is_true(a5)
guard_true(i1)
jump(a5, regs0, bytecode0, target0)
\end{verbatim}
}

@@

\begin{frame}
    \frametitle{Scaling to Large Interpreters?}
    \begin{itemize}
    \item we can apply this approach to PyPy's Python interpreter (70 KLOC)
    \item speed-ups promising (see next slide)
    \item no Python-specific bugs!
    \end{itemize}
\end{frame}

\frame[containsverbatim, plain, shrink=10]{
    \frametitle{Timings for Python Interpreter}
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
\begin{block}{Timings of \texttt{f(10000000)}}
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &1793 $\pm$ 11 &1.00\\
2 &PyPy comp'd to C, with JIT &483 $\pm$ 6 &3.71\\
3 &CPython 2.6 &1869 $\pm$ 11 & 0.96\\
4 &CPython 2.6 + Psyco 1.6 & 511 $\pm$ 7 &3.51\\\hline
\end{tabular}
\end{block}
}

@@
